Reformulating phylogenetic mixed models to improve flexibility and speed

### 2019-01-28 phyloglmm_ms.tex
stopping for now

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{color}
\usepackage{soul}
\usepackage{tabularx}
% use \xspace to allow for space after a macro as necessary
\usepackage{xspace}
\usepackage[top=1in, bottom=1in, left=1.25in, right=1.25in]{geometry}
\usepackage{pdfpages}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\usepackage{alltt}
\usepackage[sort]{natbib}
\usepackage{amsmath}
\usepackage{alltt}
\usepackage{hyperref}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage{setspace}
\usepackage{changes}
\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}

@@

Ever-increasing data collection capabilities (e.g. genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic data bases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
Researchers use these data to fit complex models describing species occurrence and traits.
In particular, phylogenetic comparative analysis (PCA) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is
becoming increasingly popular % \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e. how do numbers of citations of (whatever) increase relative to the overall size of the eco/evo literature? \ldots}
PCAs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify the phylogenetic signals in trait distributions, or both.
Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}. 
% While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}. 
Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently, they have begun to integrate evolutionary knowledge in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
While a wide range of tools is available for comparative analyses, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
In such cases, researchers must find ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
%% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
%% ord2010adaptation : used OU process with means (single obs per taxa),
%%  i.e. ignoring within-taxon variation?
%% fix me!?
% which can lead to statistical issues and erroneous conclusions
% \citep{felsenstein1985phylogenies, li2017statistical}.
% \bmb{What precise ``issues'' do you have in mind? Don't think we need to state them here, but we should know. Based on your comments, I can guess that (1) by using random slopes across taxa, Bunnefeld 2012 is neglecting more subtle variation due to differential relatedness? (2) not sure whether taking taxon-level means (Ord 2010) is actually a problem. Ignores some variation,
% but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models through random effects. 
This method allow researchers to easily incorporate various complexities without sacrificing speed.

@@

%% FIXME
The second challenge is flexibly modeling phylogenetic relations.
While it is reasonable to examine evolutionary relationships between species with their response, it is plausible the changes in effect or levels of the predictor variables have on the response variable are also phylogenetically correlated. 
%\bmb{maybe we can focus on random slopes here? Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
Several recent studies have incorporated phylogenetic signals beyond species level and looked at species response to phylogenetic signals with changes in environmental factors.
For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance via phylogenetic GLMM approach. 
The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
Table 1 provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.

@@

\begin{tabularx}{\textwidth}{|X|X|X|X|}
\hline
Model & Method & Data & Platform \\
\hline
\hline
GLM (Simple)  & Correlated residual & Single Observation & GLS, PIC \\
\hline
              & Residual    & Single Observation & Pagel's $\lambda$ \\
              & + phylo intercept &                    & Blomberg's $K$ \\ 
              &             &                    & via phylolm, GLS \\
\hline
GLMM (Complex)& Random effects & Single Observation & \pkg{pez} \\ 
              &                & Balanced Design &  \\
\hline
              &                & Unrestricted  & \pkg{lme4}, \pkg{glmmTMB} \\
\hline
Bayesian GLMM &                & Balanced Design & \pkg{MCMCglmm} \\ 
\hline        
              &                & Unrestricted   & \pkg{brms} \\
\hline
\end{tabularx}

However, in this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches.
Our new formulation is more flexible; it can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
We will compare our technique coded in R package \pkg{lme4} and \pkg{glmmTMB} with existing methods/platforms (i.e. \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) to data from simulated model that incorporates phylogenetic signal from both predictors and tips/species, as well as residual variation.
We also expect that it will be faster, because \pkg{lme4} and \pkg{glmmTMB} are optimized in performance using efficient computational linear algebra techniques. 
In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.

@@

The standard phylogenetic regression takes the form

@@

\begin{align}
\bmu & = \bX \bbeta  \label{eq:gls1} \\ 
\bY & \sim \textrm{MVN}(\bmu,\sigma^{2} \bC), \label{eq:gls2}
\end{align}
where $\bX$ is an $n \times m$ model matrix containing $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\bY$ is an $n \times 1$ response vector which is assumed to be multivariate normally distributed with mean $\bmu$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation matrix.
The PC matrix is inferred from the topology of the evolutionary tree by quantifying the degree of shared evolution between any pair of taxa \citep{garamszegi2014modern}.

@@

% More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
%\mli{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}
An alternative modelling approach is to use linear mixed effects modelling framework \citep{lynch1991methods}.
The typical linear mixed model has the form:
\begin{align}
\bY & \sim \textrm{Distr}(\bmu) \label{eq:glmm1} \\
g(\bmu) & = \bX \bbeta + \bZ \bb + \be \label{eq:glmm2} \\
\bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3} \\
\be & \sim \textrm{MVN}(0, \sigma^2_{residual} \label{eq:glmm4})
\end{align}
where $\bZ$ is an $n \times m$ model matrix for the $n$ -- dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the "G" effect) represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\be$ (sometimes referred as "R" effect) represents the residuals.
Analogously, the phylogenetic regression given by \ref{eq:gls1} can be represented in the mixed model framework by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^2_{residual} = 0$.

@@

\subsection{Reformulating phylogenetic covariance matrix}
% The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
% Phylogenetic independent contrasts is a generalization of the paired comparisons method where contrasts are taken for each bifurcation (nodes) in a phylogenetic tree. 
% Assuming that traits evolve independently in each lineage following speciation, then the trait divergences that occur at one node are independent of divergence at other nodes.

@@

An alternative approach is to model the phylogenetic correlation as a \textit{Gaussian process}. 
In particular, suppose that the evolutionary process is a Brownian motion, which means the evolution of a continuous trait is a random walk with daughter species evolving independently after speciation.  
In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. 
Thus, the evolutionary history for each species can be modeled with a sequence of independent errors with species--branch matrix $\bS$, equivalently to imposing a correlation $\bC$.
For example, as shown in figure \ref{fig:tree}, the corresponding $\bS$ is in the form:

\begin{center}
\begin{figure}[h]
  \includegraphics[scale=0.8,page=1]{./git_push/tree.pdf}
  \caption{\bmb{caption ??}}
\label{fig:tree}
\end{figure}
\end{center}

\begin{align*}
\bS = \begin{bmatrix}
\ell_1 \e_1 & \ell_2 \e_2 & 0 & 0 \\
\ell_1 \e_1 &  0 & \ell_3 \e_3 & 0 \\
0  &  0 & 0  & \ell_4 \e_4
\end{bmatrix}
\end{align*}
The error value corresponding to tip 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i$ is the square root of the branch length $L_i$ in figure \ref{fig:tree} and the $\ell_i$ are independent, homoscedastic Normal variates with zero mean and $\sigma^2$ variance (i.e. variance for tip 1 is $(\ell_1 \e_1 + \ell_2 \e_2)^2 = (\ell_1^2 + \ell_2^2)\sigma^2$, linear in branch length)

@@

The $\bS$ matrix is the matrix product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices, and a branch length vector $\bell$.

\begin{align}
\bS_{ind} = \begin{bmatrix}
1 & 1 & 0 & 0 \\ 
1 & 0 & 1 & 0 \\ 
0 & 0 & 0 & 1
\end{bmatrix} , 
\bell = \begin{bmatrix}
\ell_1 \\
\ell_2 \\
\ell_3 \\
\ell_4 
\end{bmatrix}
\label{eq:S}
\end{align}.
$\bS_{ind}$ is a binary (indicator) matrix that describes whether a particular branch occurs in the history of a focal species. 
$\bS \bS^T$ gives the variance-covariance matrix of the phylogeny.

The random-effect model matrix $\bZ$ can be decomposed into term-wise model matrix $\bZ_{i}$ as described in \citet{bates2015fitting}.
Thus, the phylogenetic correlated random-effect matrix $\bZ^{C}_{i}$ is

\begin{equation}
\bZ^{C}_{i} = (\bS^{T}_{i}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
\end{equation}
where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS_{i}$ is the $m_{i} \times b_{i}$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m_i$ indicator matrix of grouping factors; $X_{i}$ is the $n \times p_{i}$ raw random effects model matrix.

\mli{move example to the supp}
For example, using the phylogeny above (figure \ref{fig:tree}), if we begin with a raw model matrix,

@@

\begin{align}
\bZ^{C}_{i} = (\bS^{T}_{adji}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T} =
\Bigg[
\Bigg(
\begin{bmatrix}
\ell_{adj1} & \ell_{adj1}  & 0 \\
\ell_{adj2} &  0  & 0 \\
0  &  \ell_{adj3} & 0 \\
0 & 0 &  \ell_{adj4} 
\end{bmatrix}
\begin{bmatrix}
1 & 0  & 0 \\
0 & 1  & 0 \\
0 & 0  & 1  
\end{bmatrix}
\Bigg)
\ast
\begin{bmatrix}
1   & 1   & 1  \\ 
t_1 & t_2 & t_3
\end{bmatrix} 
\Bigg]^T
\\
= \begin{bmatrix}
\ell_{adj1} & \ell_{adj1}t_1 & \ell_{adj2} & \ell_{adj2}t_1 & 0 & 0 & 0 & 0 \\
\ell_{adj1} & \ell_{adj1}t_2 & 0 & 0 & \ell_{adj3} & \ell_{adj3}t_2 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & \ell_{adj4} & \ell_{adj4}t_3
\end{bmatrix}
\end{align}

@@

We generated test data based on the simple mixed model formulation (\ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3}, \ref{eq:glmm4} ) with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ = 25, 50, and 100 species.
For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
For single-site model, the full simulation model is the following:
\begin{align}
\bY & = (\beta_0 + b_{phy_{int}}) + (\beta_1 + b_{phy_{slope}}) X + e_{residual}  \label{eq:ss_sim}\\\
(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
\sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
\sigma_{phy_{int},phy_{slope}} & \sigma^2_{phy_{slope}}
\end{bmatrix} 
\Bigg) \\ 
e_{residual} & \sim \textrm{MVN} ( 0 , \Sigma) .
\end{align}
The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance, phylogenetic random slope variance and covariance between phylogenetic random slope and intercept) and observation/residual variance.  
Predictor and intercept level random effect of species are not applicable in this simulation setting with single observation per species because it will be confounded with the observational level variation (residuals).

>>>>

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{color}
\usepackage{soul}
\usepackage{tabularx}
\usepackage{multirow}
% use \xspace to allow for space after a macro as necessary
\usepackage{xspace}
\usepackage[top=1in, bottom=1in, left=1.25in, right=1.25in]{geometry}
\usepackage{pdfpages}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\usepackage{alltt}
\usepackage[sort]{natbib}
\usepackage{blkarray}
\usepackage{amsmath}
\usepackage{alltt}
\usepackage{hyperref}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage{setspace}
\usepackage{changes}
\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}

@@

Ever-increasing data collection capabilities (e.g. genomic sequencing, telemetry studies of animal behaviour, or environmental remote sensing), in combination with large-scale synthetic data bases of species occurrence and phenotypic traits, are making larger volumes of biological data available over an ever-wider taxonomic range.
Researchers use these data to fit complex models describing species occurrence and traits.
In particular, phylogenetic comparative analysis (PCA) using phylogenetic (eco-evolutionary) regression \citep{hansen2012interpreting} is becoming increasingly popular. % \bmb{might be interesting to do some bibliometry, although absolutely not necessary --- i.e. how do numbers of citations of (whatever) increase relative to the overall size of the eco/evo literature? \ldots}
PCAs explore the relationships among species traits or distributions while taking the underlying evolutionary relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify the phylogenetic signals (a measure of dependence among species responses due to their phylogenetic relationships) in trait distributions, or both.
Unlike standard statistical models, where all of the predictor variables of interest (for example, species traits and environmental factors) are directly observable,  PCAs use phylogenetic relationships to estimate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic}. 
% While evolutionary relationships are biologically important and interesting, they often get neglected in applications and multi-species analyses \citep{bunnefeld2012island, santamaria2012evolution}. 
Ecologists have used phylogenetic relationships in multi-species models \citep{garland1992procedures, freckleton2002phylogenetic, ord2010adaptation, davies2013phylogenetic}; more recently, they have begun to integrate evolutionary knowledge in applied ecological studies addressing biodiversity conservation and the effects of climate change \citep{winter2013phylogenetic, santamaria2012evolution, lankau2011incorporating, lavergne2010biodiversity, mace2008evolutionary}.
While a wide range of tools is available for comparative analyses, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
In such cases, researchers must find ways to simplify their analyses: for example, treating species effects as independent, thus neglecting phylogenetic correlations among species responses \citep{bunnefeld2012island}; ignoring degrees of relatedness and treating taxon as a strictly hierarchical description \citep{tella1999habitat}; or neglecting within-species variation \citep{ord2010adaptation}.
%% bunnefeld2012island: used taxon for random slopes but did not include phylogenetic information
%% ord2010adaptation : used OU process with means (single obs per taxa),
%%  i.e. ignoring within-taxon variation?
%% fix me!?
% which can lead to statistical issues and erroneous conclusions
% \citep{felsenstein1985phylogenies, li2017statistical}.
% \bmb{What precise ``issues'' do you have in mind? Don't think we need to state them here, but we should know. Based on your comments, I can guess that (1) by using random slopes across taxa, Bunnefeld 2012 is neglecting more subtle variation due to differential relatedness? (2) not sure whether taking taxon-level means (Ord 2010) is actually a problem. Ignores some variation,
% but it might be ignorable (cf. Murtaugh 2007 ``Simplicity and complexity in ecological data analysis'')}
In this paper, we propose an alternative method for flexibly and efficiently modeling phylogenetic relationships via mixed effect models through random effects. 
This method allow researchers to easily incorporate various complexities without sacrificing speed.

@@

%% FIXME: second sentence
The second challenge is flexibly modeling phylogenetic relations.
While it is reasonable to examine evolutionary relationships between species with their response, it is plausible the changes in effect or levels of the predictor variables have on the response variable are also phylogenetically correlated. 
%\bmb{maybe we can focus on random slopes here? Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
Several recent studies have incorporated phylogenetic signals beyond species level and looked at species response to phylogenetic signals with changes in environmental factors.
For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance via phylogenetic GLMM approach. 
The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
Table 1 provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.

@@

%% FIXME: chart looks ugly
\begin{table}[]
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{Model}                                                                                    & \textbf{Method}                                                       & \textbf{Data}                                                                & \textbf{Platform}                                                                              \\ \hline
\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)\end{tabular}}        & Correlated residual                                                   & Single Observation                                                           & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular}                                   \\ \cline{2-4} 
                                                                                                  & \begin{tabular}[c]{@{}l@{}}Residual \\ + phylo intercept\end{tabular} & Single Observation                                                           & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single Observation\\ Balanced Design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}                                                                                  \\ \hline
\multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced Design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{brms}                                                                                           \\ \hline
\end{tabular}
\end{table}

%% FIXME: second sentence
In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible.
Our new formulation can be incorporated in any framework that allows for random effect without the need to implement special correlation structures.
%% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
We will compare our technique coded in R package \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e. \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) fitting models to data from simulated model that incorporates phylogenetic signal from both predictors and tips/species, as well as residual variation.
As well as being flexible, we also expect that our approach will be fast, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
%% FIXME: move to methods/simulations
% In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
% We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.

@@

\subsection{Phylogenetic regression}

The standard phylogenetic regression can be formulated as

@@

\begin{align}
\bmu & = \bX \bbeta  \label{eq:gls1} \\ 
\bY & \sim \textrm{MVN}(\bmu,\sigma^{2} \bC), \label{eq:gls2}
\end{align}
where $\bX$ is an $n \times m$ model matrix, describing $n$ observations of $m$ predictor variables (phenotypic traits or environmental variables, typically including an intercept column of ones); $\bbeta$ is an $m$-vector of coefficients; $\bY$ is an $n \times 1$ response vector which is assumed to be multivariate normally distributed with mean $\bmu$ and variance-covariance matrix given by $\sigma^{2} \bC$ where $\bC$ is a $n \times n$ phylogenetic correlation matrix.
The PC matrix is inferred from the topology of the evolutionary tree by quantifying the degree of shared evolution between any pair of taxa \citep{garamszegi2014modern}.

@@

% More recently, researchers use linear mixed model and generalized linear mixed model framework to model complex systems with phylogenetic structures.
%\mli{Why? Data type, interactions, random effects, etc... Need to really explain random effects here. Alternatively, we can drop this line and write this...}

\subsection{Phylogenetic generalized linear mixed model}
Alternatively, one can use the generalized linear mixed effects modelling (GLMM) framework \citep{lynch1991methods}.
The classic generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorportate more than one kind of variability through random effects and for non-gaussian responses.
The typical GLMM has the form:
\begin{align}
\bY & \sim \mathbf{f}(\bmu,\bSigma) \label{eq:glmm1} \\
\bmu & = \bX \bbeta + \bZ \bb  \label{eq:glmm2} \\
\bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3} \\
\end{align}
where $\bZ$ is an $n \times m$ model matrix for the $n$ -- dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the "G-side" effect) represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\be$ (sometimes referred as "R" effect) represents the residuals. %% FIXME: see basis-fn paper
Analogously, the phylogenetic regression given by (\ref{eq:gls1}) can be represented in the mixed model framework by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^2_{residual} = 0$.

There are serveral advantages in modelling phylogenetic correlated traits in the mixed model framework.
First, random intercepts can allow the response trait to vary across groups/levels within grouping predictors (e.g. patch, site, experiment, and etc). 
Second, it can also allow the response (trait or distribution) to vary according to the phylogenetic relationships across species (i.e. similar species will have similiar responses), and between species (tip variation).
The third type of variation that is often neglected or rarely considered are random slopes.
Random slopes can allow fixed effects (size of the effect in continuous predictor variables) to vary for each species. 
Analogous to modelling the repsonse dependency via phylogenetic relationship among species, phylogenetic random slopes can allow fixed effects to account for the phylogenetic relations (i.e. similar species will have similar size of the effects on the response than distinct pairs).

@@

\subsection{Reformulating the phylogenetic covariance matrix}
% The standard problem of phylogenetic comparative methods is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree.
% Phylogenetic independent contrasts is a generalization of the paired comparisons method where contrasts are taken for each bifurcation (nodes) in a phylogenetic tree. 
% Assuming that traits evolve independently in each lineage following speciation, then the trait divergences that occur at one node are independent of divergence at other nodes.

@@

An alternative approach is to model the phylogenetic correlation as a \textit{Gaussian process}. 
In particular, suppose that the evolutionary process is a Brownian motion, which means that evolution of a continuous trait evolve independently, following a standard Brownian motion, along each branch of the phylogeny.
In this case, the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
Thus, modeling the evolutionary history of each species with a sequence of independent errors with species--branch matrix $\bS$, is equivalent to imposing a correlation $\bC$.
For example, for the phylogeny in figure \ref{fig:tree}, the corresponding $\bS$ takes the form:

%% FIXME: need top/side labels
\begin{center}
\begin{figure}[h]
  \includegraphics[scale=0.8,page=1]{./git_push/tree.pdf}
  \caption{Three species phylogenetic tree.}
\label{fig:tree}
\end{figure}
\end{center}

\[
\begin{blockarray}{ccccc}
 & L_1 & L_2 & L_3 & L_4  \\
\begin{block}{c(cccc)}
  t_1 & \ell_1 \e_1 & \ell_2 \e_2 & 0           & 0 \\
  t_2 & \ell_1 \e_1 &  0          & \ell_3 \e_3 & 0 \\
  t_3 & 0           &  0          & 0           & \ell_4 \e_4 \\
\end{block}
\end{blockarray}
 \]

%% FIXME: double check this
The phylogenetic variability corresponding to species 1 is $\ell_1 \e_1 + \ell_2 \e_2$, where $\ell_i = \sqrt{L_i}$, the square root of the branch length $L_i$ in figure \ref{fig:tree} and the $\ell_i$ are independent, homoscedastic Normal variates with zero mean and $\sigma^2$ variance (i.e. the variance for species 1 is $(\ell_1 \e_1 + \ell_2 \e_2)^2 = (\ell_1^2 + \ell_2^2)\sigma^2$, linear in branch length). 
%% FIXME: see BB's cmt

@@

The $\bS$ matrix is the product of an $n \times b$ indicator matrix $\bS_{ind}$ of branch indices and a branch length vector $\bell$.

\begin{align}
\bS_{ind} = \begin{bmatrix}
1 & 1 & 0 & 0 \\ 
1 & 0 & 1 & 0 \\ 
0 & 0 & 0 & 1
\end{bmatrix} , 
\bell = \begin{bmatrix}
\ell_1 \\
\ell_2 \\
\ell_3 \\
\ell_4 
\end{bmatrix} .
\label{eq:S}
\end{align}
$\bS_{ind}$ is a binary (indicator) matrix that describes whether a particular branch occurs in the history of a focal species. 
$\bS \bS^T$ gives the variance-covariance matrix of the phylogeny.

The random-effect model matrix $\bZ$ can be decomposed into term-wise model matrices $\bZ_{i}$ as described in \citet{bates2015fitting}.
Analogous to the procedure described in \citet{bates2015fitting}, the phylogenetic correlated random-effect matrix $\bZ^{C}_{i}$ is

\begin{equation}
\bZ^{C}_{i} = (\bS^{T}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T}, \label{eq:ZC}
\end{equation}
where $\ast$ is the Khatri-Rao product \citep{khatri1968solutions}, $\bS$ is the $m \times b$ species--branch matrix; $\bJ_{i}$ is the $n_i \times m$ indicator matrix of grouping factors; and $\bX_{i}$ is the $n \times p_{i}$ raw random effects model matrix.

For example, using the phylogeny above (figure \ref{fig:tree}), if we begin with a raw model matrix corresponding to a random-slope model,

@@

\begin{align}
\bZ^{C}_{i} = (\bS^{T}_{adji}\bJ^{T}_{i} \ast \bX^{T}_{i})^{T} =
\left[
\left(
\begin{bmatrix}
\ell_{adj1} & \ell_{adj1}  & 0 \\
\ell_{adj2} &  0  & 0 \\
0  &  \ell_{adj3} & 0 \\
0 & 0 &  \ell_{adj4} 
\end{bmatrix}
\begin{bmatrix}
1 & 0  & 0 \\
0 & 1  & 0 \\
0 & 0  & 1  
\end{bmatrix}
\right)
\ast
\begin{bmatrix}
1   & 1   & 1  \\ 
t_1 & t_2 & t_3
\end{bmatrix} 
\right]^T
\\
= \begin{bmatrix}
\ell_{adj1} & \ell_{adj1}t_1 & \ell_{adj2} & \ell_{adj2}t_1 & 0 & 0 & 0 & 0 \\
\ell_{adj1} & \ell_{adj1}t_2 & 0 & 0 & \ell_{adj3} & \ell_{adj3}t_2 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & \ell_{adj4} & \ell_{adj4}t_3
\end{bmatrix}.
\end{align}

@@

We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3}, \ref{eq:glmm4} ) with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ = 25, 50, and 100 species.
For simplicity, the response variable $\bY$ is conditional normally distributed (i.e. gaussian distribution with an idenitiy link function) , which is a linear mixed effect model. 
For the first set of simulation, we simulate one observation per species.
% For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
The full simulation model is the following:
\begin{align}
\bY & = (\beta_0 + b_{phy_{int}}) + (\beta_1 + b_{phy_{slope}}) X + e_{residual}  \label{eq:ss_sim}\\\
(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
\sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
\sigma_{phy_{int},phy_{slope}} & \sigma^2_{phy_{slope}}
\end{bmatrix} 
\right) \\ 
e_{residual} & \sim \textrm{MVN} ( 0 , \Sigma) .
\end{align}
The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance, phylogenetic random slope variance and covariance between phylogenetic random slope and intercept) and residual variance.  
The covariance between phylognetic random intercept and slope measures the correlation of phylogenetic dependency variability of the size in effect and outcome.
\mli{con't here!}
Predictor and intercept level random effect of species are not applicable in this simulation setting with single observation per species because it will be confounded with the observational level variation (residuals).
